---
title: "RXposé"
author: "Aaron"
date: "`r Sys.Date()`"
knit: (function(input, encoding) {rmarkdown::render(input, output_file = 'README')})
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Rxposé: An Analysis of Performance-Enhancing Drug Use in the Sport of Weightlifting
Note : This portfolio project presents an opportunity to practice SQL skills, as well as learning how to link a database and run multiple languages within an Rmarkdown. There are places where it may be simpler to use only R, but it is a valuable learning experience, regardless.

## Background
Weightlifting has been one of the sports with the highest (recorded) prevalence of drug use. Rumors are always circling, and considerable evidence has shown that covert as well as state-sponsored doping rings are commonplace. Many people believe that the number of "clean" athletes at the elite level is extremely low, possibly as low as zero.

While the analysis of the effectiveness of individual tests is fascinating and important, the initial scope of this project is much broader. The question at the center of this study is: 

Has drug testing made an impact on the performance of athletes in the sport?

To answer this we can compare two populations, athletes who have tested positive for performance-enhancing drugs (PEDs) against those who have not. This tests a central assumption about drug testing, i.e. PEDs increase performance. If the positive-testing athletes (sometimes referred to as "popped") do not outperform the negative-testing athletes, the truth of this assumptions falls must be questioned.

A quick primer on the sport of weightlifting: as sports go, it is relatively simple. Whoever lifts the most weight wins. Athletes are separated by assigned sex (male/female) and weight class. Weight class is a way of separating body types, as being larger and having more muscle allows people to typically lift more weight. Usually, the athletes are split into weight classes that are about 7kg apart. This means that all the athletes competing against one another weigh within 7kg (15 lbs) of each other. This allows broader participation in the sport, since it allows people who are not naturally large to compete. The standard modern convention is to identify a weight class by the maximum allowable bodyweight in that weight class. For example, the M 102kg category means that any man weighing less than 102kg is eligible to compete. (Note: athletes may always compete in a heavier category than their bodyweight, but it is almost never advantageous to do so, as the other competitors will be larger and stronger.)

In order to compare all athlete, the [Sinclair Coefficient](https://iwf.sport/weightlifting_/sinclair-coefficient/) will be employed. This will allow us to compare performances of athletes across weight classes and get a better understanding of the influence of drug testing on the sport as a whole.

We should note that the subjectivity of "talent" will always be part of this discussion. This conversation is inextricably linked with drug-testing, as genetics affect athletic performance, drug response, and even our ability to screen for drug use.

Additional topics for potential future study:

* Prevalence of use by country, region, gender, etc.

* Life-time competition bans vs. temporary bans

* When in the athlete's career a ban occurred.

## Data

One of the advantages of studying sports is the prevalence of well-kept records. Weightlifting in particular is convenient due the the simple measurement of performance-- more weight lifted (within a weight class) is a proportionally better performance.

For Olympic competitions, [this dataset](https://www.kaggle.com/datasets/piterfm/olympic-games-medals-19862018) was immensely helpful in providing performance results. A huge thank-you to Petro for providing it, and keeping it current.

At least for now, we will restrict our analysis to Olympic competition. While this will select for the highest-performing athletes, the Olympics are highly visible and widely regarded. It makes sense from a public perception standpoint to focus here.

First, the data is loaded into a SQL databsae using RSQLite. There are limitaions with this package, but it is a good opportunity to practice queries.

```{r}
## Packages for this analysis
library(tidyverse)
library(RSQLite)
library(DBI)
library(dbplyr)
getwd()
setwd("~/Documents/Data Analytics Projects/Rxpose/Kaggle Olympics Data")
olympics_data_raw <- read_csv("olympic_results.csv")

con <- DBI::dbConnect(RSQLite::SQLite(), 
        dbname = "olympics_db.sqlite")

dbWriteTable(con, 
             "all_olympics_results", 
             olympics_data_raw, 
             overwrite = TRUE)
```

Only weightlifting data was extracted for analyis, and only from 1976 and later. This is due to two reasons. First, there was a [rule shift](https://en.wikipedia.org/wiki/Clean_and_press) after 1972 that greatly affected the total weight lifted. Second, anabolic steroids were not banned until 1975. [source](https://www.acog.org/clinical/clinical-guidance/committee-opinion/articles/2011/04/performance-enhancing-anabolic-steroid-abuse-in-women#:~:text=Anabolic%20steroids%20were%20first%20discovered,performance%20and%20enhance%20cosmetic%20appearance.)

Below is the code that was run in SQLite within the markdown file. This filters for only weightlifting results from the correct years, and it extracts and casts the competition year and the weight lifted as more usable data types. It also removes any null or 0 values from the total_kg column, as those values will not be useful for comparison.

```{sql connection=con}
CREATE TABLE IF NOT EXISTS weightlifting_results AS
SELECT event_title, rank_position, country_3_letter_code, athlete_full_name,
  value_type,
  CAST(value_unit AS decimal) AS total_kg,
  CAST(SUBSTR(slug_game, -4) AS INT) AS year
FROM
  all_olympics_results
WHERE
  discipline_title = "Weightlifting" 
  AND
  value_type IS NOT NULL
  AND
  total_kg IS NOT NULL
  AND
  total_kg >0
  AND
  year > 1972

-- Why can I not use "year" as a variable here? Add it to the table with a join?
-- Works here, just not in BigQuery
-- CAST(SUBSTR(slug_game, 4) AS INT) >1972 for use in BigQuery
```


#### Separating Sex and Weight Class
First, we will add columns to designate men's and women's events, as well as for weightclass.


```{sql connection=con}
ALTER TABLE weightlifting_results 
  ADD sex NOT NULL default 'Men'
```
Note: this is not to say that men's sports are the default, but it is easier to use REGEX to select for the string "women" rather than "men."

```{sql connection=con}
UPDATE weightlifting_results
  SET sex = CASE 
    WHEN (event_title LIKE '%women%') THEN 'F'
    ELSE 'M' END

```

```{sql connection=con, eval = FALSE}
-- just a check for the sex column
SELECT DISTINCT sex, country_3_letter_code 
FROM weightlifting_results
LIMIT 20
```

At this point, we need to clean the disastrous formatting that is the event_title column. The strings are not consistent, and we need to extract a usable number from them. Weight classes are determined by the maximum allowable weight in each category, so we will will need to extract the largest number present in each string. Usually, there is only one number, but this column sometimes gives a range of maximum and minimum weights when we only want the larger value. From the table below, we can see that there are even some misprints in the numbers themselves (e.g. "825" kg should be "82.5").

This step of cleaning required r, as SQLite does not support REGEX. Data were queried, cleaned, then reuploaded to replace the original table.

```{r}
weightlifting_results <- 
  dbGetQuery(con, "SELECT * FROM weightlifting_results")


weightlifting_results_after_regex_extraction <-
weightlifting_results %>%
  mutate(weight_class_kg = as.numeric(str_extract(event_title, "([0-9.])+(?=kg)")
  ))

unique(sort(weightlifting_results_after_regex_extraction$weight_class_kg)) ## check for necessary changes

## Since 825kg is a nonsensical number in weightlifting, we can universally replace 825 with 82.5.

weightlifting_results_after_regex_extraction[weightlifting_results_after_regex_extraction == 825] <- 82.5
unique(sort(weightlifting_results_after_regex_extraction$weight_class_kg))

```
This code conflates the weight classes for the heaviest and second-heaviest categoies, but that will be sorted out with SQL. This resulting dataframe can be uploaded to our database.

```{r}
dbWriteTable(con, 
             "weightlifting_results", 
             weightlifting_results_after_regex_extraction, 
             overwrite = TRUE)
```

```{sql connection=con}
-- test to make sure the new table exists and has correct data
SELECT *
FROM weightlifting_results
LIMIT
20
```



#### Sinclair considerations
Sinclair calculations are not perfect, as different weight classes are essentially playing different games. For example, the goal of a superheavyweight lifter is to get as big and strong as possible to lift maximum weight, while lighter lifters need to get as big and strong as possible within a specified boundary. Obviously, different body types will self-select into their most competitive classes over the course of time. For this study, the maximum allowable weight for each category will be used. For non-superheavy weight athletes, this approximation is very good, as it is in their best interest to be as heavy (read muscular, strong) as possible while remaining in their category. The Sinclair method includes a "maximum weight" that is used for these corrections, and that is the bodyweight used in this study for all supers. While this may lower the Sinclair totals for superheavyweight athletes, since this study is not comparing individual lifters, this is an acceptable apporoximation. As a second approximation, the calculated values for the current Olympic cycle will be used. This will greatly simplify calcuations, and it will allow modern scaling of historical athletes while crediting them with their bodyweight at the time.

For this analysis, credited bodyweights for male and female superheavyweights were taken from the [IWF website](https://iwf.sport/wp-content/uploads/downloads/2023/05/2021-Sinclair_Coefficients.pdf).
```{sql connection=con}
UPDATE weightlifting_results
  SET weight_class_kg = 
    CASE 
      WHEN (sex = 'M' AND event_title LIKE '%+%') THEN 193.609
      WHEN (sex = 'M' AND event_title LIKE '%super%') THEN 193.609
      WHEN (sex = 'F' AND event_title LIKE '%+%') THEN 153.757
      WHEN (sex = 'F' AND event_title LIKE '%super%') THEN 153.757
      ELSE weight_class_kg
  END
  
-- need both "super" and "+" to be accounted for
-- [WHEN (event_title LIKE '%+%') THEN 1000 ELSE (1)]
-- couldn't get the OR function to work inside LIKE function
        
```

```{sql connection=con}
SELECT DISTINCT weight_class_kg, event_title
FROM weightlifting_results
ORDER BY weight_class_kg DESC

```

#### Adding Sinclair to the table
Finally, we get to add the primary value for analysis to this table.

```{sql connection=con}
ALTER TABLE weightlifting_results 
  ADD sinclair NOT NULL default 0
```

```{sql connection=con}
UPDATE weightlifting_results
  SET sinclair = CASE
    WHEN (sex = 'F')
    THEN  total_kg* POWER(10, 0.787004341*POWER(LOG10(weight_class_kg/153.757),2))
    ELSE  total_kg* POWER(10, 0.722762521*POWER(LOG10(weight_class_kg/193.609),2))
    END

```

```{sql connection=con}
SELECT DISTINCT weight_class_kg, sinclair, athlete_full_name
FROM weightlifting_results
WHERE sex = 'F'
ORDER BY sinclair DESC
LIMIT 30

```
While not the correct Sincalir for each individual athlete, this will serve well enough for the purposes of this study. This list is absolutley not intended to rank athletes' performances. I would like to apologize especially to Lasha Talakhadze, who as the best superheavyweight in history appears at #28 on this ranking. Using historical Sinclair calculations, he appears at #2 behind Naim Süleymanoglu, but I still have faith that Lasha will hit the legendary 500kg total before he retires.

It is interesting to note that two female superheavyweights appear in the top 10 in this ranking. I will not analyze that too much here, but there is likely a huge amount of influence on this calcultion by representation in the sport. The number of male athletes (historically) is far greater, and it would be wonderful to see how more female representation would change the sport.

As the last step in data cleaning and preparation, we need to add a table of lifters who have served drug bans. A list (if not a complete one) exists at [this link](https://en.wikipedia.org/wiki/Category:Doping_cases_in_weightlifting). After [stripping the html](https://www.striphtml.com/), I added the list to my [spreadsheet](https://docs.google.com/spreadsheets/d/1vWlYv58PvmJE1VBEh5-dOVboG6D-HxOqWq_zTylOWC0/edit?usp=sharing).

Creating the new table.
```{r}
setwd("~/Documents/Data Analytics Projects/Rxpose/Kaggle Olympics Data")
weightlifters_with_doping_violations_raw <- read_csv("weightlifters_with_doping_violations.csv")
str(weightlifters_with_doping_violations_raw)
```
This table will also require some in-depth cleaning. The table includes parentheses (left over from the online source), and there are many spellings of names with non-English characters. Matching names and successfully executing a join with the previous table will require REGEX, so this portion of cleaning will necessitate R.

```{r}
## This code removes the parentheses from the strings in the athlete's names and creates a new column for the cleaned names

weightlifters_with_doping_violations_no_parentheses <-
weightlifters_with_doping_violations_raw %>%
    mutate(weightlifters_with_doping_violations_no_parentheses = gsub("\\(.*$","",weightlifters_with_doping_violations_raw$weightlifters_with_doping_violations))
## Note to self on REGEX-- the double backslash is important due to "\" and "(" both being escape operators.


# Creates a table with only one column
weightlifters_with_doping_violations_no_parentheses_v_2 <- weightlifters_with_doping_violations_no_parentheses[2]

```

The final step in analysis is to join the table of weightlifters with violations with the table of all lifters to separate the two populations. In the table weightlifters_with_doping_violations_no_parentheses, there are many letters considered "special characters" in the English alphabet. There also appear to be extraneous capital letters, possibly from odd formatting from other alphabets. For simplicity, we can match special characters to the English character on which they are based. For example, "î","ï", "í", and "ì" would all match "i." We check the encoding and get some bad news.

```{r}
Encoding(weightlifters_with_doping_violations_no_parentheses_v_2$weightlifters_with_doping_violations_no_parentheses)

# Lots of code below that I don't want to delete yet in case I need it later...

#weightlifters_with_doping_violations_reduced <- iconv(weightlifters_with_doping_violations_no_parentheses_v_2$weightlifters_with_doping_violations_no_parentheses,from="UTF-8",to="ASCII//TRANSLIT")
#Encoding(weightlifters_with_doping_violations_reduced)

#weightlifters_with_doping_violations_no_parentheses_v_2$weightlifters_with_doping_violations_no_parentheses = stri_trans_general(str = weightlifters_with_doping_violations_no_parentheses_v_2$weightlifters_with_doping_violations_no_parentheses, id = "Latin-ASCII")
#Encoding(weightlifters_with_doping_violations_no_parentheses_v_2$weightlifters_with_doping_violations_no_parentheses)
#view(weightlifters_with_doping_violations_no_parentheses_v_2)
#library(stringi)
#weightlifters_with_doping_violations_no_parentheses_v_2[, "weightlifters_with_doping_violations_no_parentheses" = stri_trans_general(str = weightlifters_with_doping_violations_no_parentheses_v_2, id = "Latin-ASCII")]
```
Unknown encoding will be difficult to "translate." 


## Analysis










##### This was method number 1--This work was improved upon via other methods, but it will be remain here for reference at least until the first iteration of the project is complete.

To begin the cleaning process, I created a new table with the best performances by each athlete. (This may be an issue, as it might eliminate performances of athletes that rank better by Sinclair if they moved up a category and underperformed.)
```{r}
##SELECT MAX(total_kg) AS best_total_kg, athlete_full_name,
##FROM `weightliftinganalysis.Olympics_Results_86_to_20.All_weightlifting_performances_##      post_1976`
##WHERE total_kg IS NOT NULL
##GROUP BY
##  athlete_full_name

```
Then, I joined this table back to the original to have the full table with only the best performance of each athlete. A big thank you to Bill Karwin for his [answer](https://stackoverflow.com/questions/121387/fetch-the-rows-which-have-the-max-value-for-a-column-for-each-distinct-value-of) on how to do this efficiently.
```{r}
##SELECT *
##FROM`weightliftinganalysis.Olympics_Results_86_to_20.All_weightlifting_performances_p## ost_1976`
##LEFT OUTER JOIN
##  `weightliftinganalysis.Olympics_Results_86_to_20.Only_nonzero_performances_post_1976`
##  ON (`weightliftinganalysis.Olympics_Results_86_to_20.Only_nonzero_performances_post_1976`.athlete_full_name = `weightliftinganalysis.Olympics_Results_86_to_20.All_weightlifting_performances_post_1976`.athlete_full_name
##    AND
##    `weightliftinganalysis.Olympics_Results_86_to_20.Only_nonzero_performances_post_1976`.best_total_kg >
##    `weightliftinganalysis.Olympics_Results_86_to_20.All_weightlifting_performances_post_1976`.total_kg)
##  WHERE
##    total_kg IS NOT NULL AND
##    best_total_kg IS NULL
```

This almost worked perfectly, but it still will repeat athletes if they had exactly the same totals in multiple competitions. This is an easier problem to solve in a spreadsheet, so at this point I moved the .csv file to that software. With less than 2000 rows, I am not concerned about performance.

### Spreadsheet cleaning and analysis
#### [This link](https://docs.google.com/spreadsheets/d/1vWlYv58PvmJE1VBEh5-dOVboG6D-HxOqWq_zTylOWC0/edit?usp=sharing) will show the spreadsheet in its current form.

As mentioned before, there are a few names which repeat, but only 14. These can be sorted out without too much difficulty. In fact, it might make sense to leave them in and credit the athlete with the larger Sinclair.

The larger problem is that of weight classes. Labelled "event_title" in this dataset, there is little to no consistency in how these data were entered. It is necessary to have a consistent format, as weight class and year will determine Sinclair calculations.


  

```{sql connection=con, eval = FALSE}
DROP TABLE IF EXISTS weightlifting_results

```

